Red Wine quality exploration by Zhengda He
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages 
library(ggplot2)
library(GGally)
library(gridExtra)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
wine = read.csv('wineQualityReds.csv')
```

This report explores the red wine quality dataset, which public available for research.

The goal of this report is to explore Which chemical properties influence the quality of red wines.

# Univariate Plots Section


```{r echo=FALSE, Univariate_Plots}

str(wine)
summary(wine)
```
### Quality

```{r echo=FALSE}

# first histogram of quality
ggplot(aes(x = quality), data = wine)  + 
geom_bar(aes(y = (..count..)/sum(..count..))) + 
ggtitle("Plot of quality distribution") + 
ylab('percentage distribution')
print('summary of wine quality')
summary(wine$quality)
print('distribution of wine quality')
table(wine$quality)

```
From the plot and statistical data we could see that over 80% sample of our quality level falls in 5 or 6. Nearly 95% falls in level 5,6 or 7.
```{r echo=FALSE}
plot.hist <- function(x,bw){
    ggplot(wine, aes_string(x = x)) +
    geom_histogram(binwidth = bw)
}
```
### Fix acid
```{r echo=FALSE}
plot.hist('fixed.acidity',0.1) +  
scale_x_continuous(lim = c(4, 16)) + 
ggtitle('Fix acid distribution')
summary(wine$fixed.acidity)
```
Fixed acidity is the measurement of tartaric acid (g/dm^3) in the wine.

Hist plot shows that this variable is left skewed

### Volatile acidity"
```{r echo=FALSE}

plot.hist('volatile.acidity',0.02) +
ggtitle('Volatile acid distribution')
summary(wine$volatile.acidity)

```
Volatile acidity is the measurement of acetic  acid (g/dm^3) in the wine.

This hist plot also shows that it is left skewed, with tail on the right.

### Citric acid
```{r echo=FALSE}

plot.hist('citric.acid',0.01)+ 
ggtitle('Citric acid distribution')
summary(wine$citric.acid)
print(paste("number of 0 citric acid: ",
            length(subset(wine,wine$citric.acid == 0)$citric.acid)))
print(paste("number of 0 citric acid: ",
            length(subset(wine,wine$citric.acid == 1)$citric.acid)))

```
citric acid is the measurement of citric acid (g/dm^3) in the wine. 

From the plot we could clearly see there is a spike on the left of the plot, which indicates that there are over 100 sample in this dataset has 0 citric acid measurement.From console output we know that there is 1 sample has citric acid 1 and 132 sample with citric acid value 0.

### Residual sugar
```{r echo=FALSE}

plot.hist('residual.sugar',0.1)+ 
ggtitle('Residual sugar distribution')
summary(wine$residual.sugar)
IRQ <- 2.60 - 1.90
print(paste("3rd quatile + 1.5*IRQ = ", 2.6 + 1.5*IRQ)) 
print(paste("number of point could be considered as outlier:",
      length(subset(wine,residual.sugar >= 2.6 + 1.5*IRQ)$residual.sugar)))

```
Residual sugar is the amount of sugar remaining after fermentation stops in the wine, it measures in (g/ dm^3)\

From the plot we could see most of our sample falls in the low sugar region.

Through statistical analysis we found that IRQ = 0.7,while 3rd quatile value is 2.6, any point beyond the value (2.6+0.7*1.5) = 3.65 could be considered as outlier. 

Thus we could plot this again with proper range(1-4)
```{r echo=FALSE}

plot.hist('residual.sugar',0.1) + 
scale_x_continuous(lim = c(1,4)) + 
ggtitle('Residual sugar distribution')
```

This looks much better now.

### chlorides
```{r echo=FALSE}

plot.hist('chlorides',0.005) + 
ggtitle('chlorides distribution')
summary(wine$chlorides)
IRQ <- 0.09 - 0.07
print(paste("3rd quatile + 1.5*IRQ = ", 0.09 + 1.5*IRQ)) 
print(paste("number of point could be considered as outlier:",
    length(subset(wine,residual.sugar >= 0.09 + 1.5*IRQ)$residual.sugar)))
```
chlorides is the amount of salt in the wine, measured in (g/dm^3)
Similar to the residual sugar plot, almost all of the sample falls in the left of the plot while only few outlier in the right.
Through statistical analysis we found that IRQ = 0.02,while 3rd quatile value is 0.09, any point beyond the value (0.09+0.0.02*1.5) = 0.11 could be considered as outlier. 
Thus we could plot this again with proper range(0-0.02)
```{r echo=FALSE}

plot.hist('chlorides',0.001) + 
scale_x_continuous(lim = c(0,0.2)) + 
ggtitle('chlorides distribution')
```

This looks much better now  

### Free sulfur dioxide
```{r echo=FALSE}

p2 <- plot.hist('free.sulfur.dioxide',0.05) + 
ggtitle('Free sulfur dioxide distribution log10') + 
scale_x_log10()
p1 <- plot.hist('free.sulfur.dioxide',1) + 
ggtitle('Free sulfur dioxide distribution')
grid.arrange(p1, p2, ncol = 2)
summary(wine$free.sulfur.dioxide)
```

Free sulfur dioxide is the free form of SO2 exists in equilibrium between molecular SO2 (as a dissolved gas) and bisulfite ion. 
The left plot is original distributuib plot, we could easily see that it is left skewed, with long tail on the right. 
The right plot is the original plot with log10 on the x aixs, on this plot the long tail disappeared.

### Total sulfur dioxide
```{r echo=FALSE}

p2 <- plot.hist('total.sulfur.dioxide',0.05) + 
ggtitle('Total sulfur dioxide distribution log10') + 
scale_x_log10()
p1 <-plot.hist('total.sulfur.dioxide',2) + 
ggtitle('Total sulfur dioxide distribution')
grid.arrange(p1, p2, ncol = 2)
summary(wine$total.sulfur.dioxide)
```

Total sulfur dioxide is amount of free and bound forms of S02.

Similar to free sulfur dioxide, total sulfur dioxide is also left skewed with long tail on the right. Take log10 on the x aixs make it more normal distribution.

### Density
```{r echo=FALSE}
plot.hist('density',0.0001) + ggtitle('density distribution')
summary(wine$density)
```
Density represents the density of the wine measure in (g / cm^3).  
The density variable has a comparatively normal distribution, centered on 1g/cm^3(which is the density of water)

### pH

```{r echo=FALSE}
plot.hist('pH',0.01)  + ggtitle('pH distribution')
summary(wine$pH)
```

pH describes how acidic or basic a wine is on a scale from 0 (very acidic) to 14 (very basic).

Similar to density, pH also has a faily normal distribution.

### sulphates

```{r echo=FALSE}
p2 <- plot.hist('sulphates',0.01) + 
  ggtitle('sulphates distribution log10') + 
  scale_x_log10()
p1 <- plot.hist('sulphates',0.01) + 
  ggtitle('sulphates distribution')
grid.arrange(p1, p2, ncol = 2)
summary(wine$sulphates)
```

sulphates is a wine additive which can contribute to sulfur dioxide gas (S02) levels, wich acts as an antimicrobial and antioxidant.

Similar to both sulfur dioxide in our dataset,sulphates plot (the left one) is also left skewed, with long tail on the right.  

Take log10 on the x aixs make it have more normal distribution.

### alcohol

```{r echo=FALSE}
plot.hist('alcohol',0.2) + ggtitle('alcohol distribution')
summary(wine$alcohol)
```

Alcohol refers to the percent alcohol content of the wine, with % by volume.

Alcohol distribution is left skewed, with a spike around 9.5, and decrease gradually to 0 on 14.

# Univariate Analysis

### What is the structure of your dataset?
This is a tidy dataset, with 1599 observation on red wine quality and its chemical characteristics.

### What is/are the main feature(s) of interest in your dataset?

The main feature of interest in the dataset should be the quality of red wine, the goal of this exploration is to figure out how other feature affect the quality feature.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

From the description in the dataset, it seems all 11 feature could affect the taste of the wine,thus influence the quality of the wine.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

I did not make any modification on the dataset, since this is a pretty tidy dataset, with no missing value, and all the feartures seems to independent to each other.

# Bivariate Plots Section


```{r echo=FALSE, Bivariate_Plots}
# generate a bivariate grid using GGally
ggpairs(wine[,2:13]) + ggtitle("pair plot in red wine dataset")
```

From the ggpair plot we could see that there are clearly some feature has correlation with quality.  

Alcohol seems affect the most(correlation 0.476), followed by volatile acidity(correlation -0.391). Although affect the most, the correlation score is not high(beyong 0.7) which indicates that it is not strong correlated. 

Then is slightly low correlated feature, includes:sulphates( correlation 0.251) and citric acid(correlation 0.226) 
Other features, which below correlation +-0.2, are considered no correlation with quality.

### Alcohol Vs Quality

```{r echo=FALSE}

p1<- ggplot(aes(x = alcohol, y = quality), data = wine) +
  geom_jitter(alpha = 0.1)  + 
  geom_smooth(method = 'lm') + 
  ggtitle("alcohol vs quality scatter plot with \n trend line")
p2<-ggplot(aes(x = factor(quality), y = alcohol), data = wine) +
  geom_boxplot()  +
  xlab("quality") + ggtitle("alcohol vs quality box plot")

grid.arrange(p1,p2,ncol = 2)
```

Scatter plot with smooth line above shows the relationship between quality and alcohol, who has the most correlation with quality. 

The smooth line indicates the overall correlation of quality and alcohol.

Box plot above also reveals the overall trends: Higher quality the wine has, higher alcohol it contains. 


### volatile acidity Vs Quality

```{r echo=FALSE}
# generate a bivariate grid using GGally
p1<- ggplot(aes(x = volatile.acidity , y = quality), data = wine) +
  geom_jitter(alpha = 0.1)  + 
  geom_smooth(method = 'lm') +  
  ggtitle("volatile acidity vs quality scatter plot with trend line")

p2<- ggplot(aes(x = factor(quality), y = volatile.acidity), data = wine) +
  geom_boxplot()  +
  xlab("quality") + ggtitle("volatile acidity vs quality box plot")
grid.arrange(p1,p2,ncol = 2)
```



Similar to plot in quality vs alcohol, the the points from scatter point could help visualizing the distribution of our samples. Smooth line could be viewed as trend line, help visualizing the correlation of the two features. Box plot could help visualizing statistic data in different group.

In the plot above, we could see that volatile acidity and quality are negatively correlated, which demostrated the correlation score.

### sulphates  Vs Quality and Citric acid vs Quality

```{r echo=FALSE}
# generate a bivariate grid using GGally
p1 <-ggplot(aes(x = sulphates , y = quality), data = wine) +
  geom_jitter()  + 
  geom_smooth(method = 'lm') +  ggtitle("sulphates vs quality scatter plot \n with trend line")
p2 <- ggplot(aes(x = factor(quality), y = sulphates ), data = wine) +
  geom_boxplot()  +
  xlab("quality") + ggtitle("sulphates vs quality box plot")
p3 <-ggplot(aes(x = citric.acid , y = quality), data = wine) +
  geom_jitter()  + 
  geom_smooth(method = 'lm')+  ggtitle("citric acid vs quality scatter plot \n with trend line")
p4 <- ggplot(aes(x = factor(quality), y = citric.acid ), data = wine) +
  geom_boxplot()  +
  xlab("quality")+ ggtitle("citric acid vs quality box plot")
grid.arrange(p1,p2,p3, p4, ncol = 2) 

```

These are both low correlation feature with quality.  
From the plot we could see they are both positively correlated with quality, but the slope of trend line will drop down, as the correation decreases.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?
Quality tend to change its value with the change of other features, but not significant in our sample dataset.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

From ggpair plot we could find some strong correlations, such as density vs fix acidity(0.668), pH vs fix acidity(-0.683),total sulfur dioxide vs free sulful dioxide(0.668), and all of these seems to make sense.

### What was the strongest relationship you found?
Alcohol, with correlation score 0.476

# Multivariate Plots Section
```{r echo=FALSE}
plot.multi <- function(x,y){
    ggplot(wine, aes_string(x = x,y = y)) +
    geom_jitter(aes(color = factor(quality) ), alpha = 0.5)+
    scale_color_brewer(type = 'seq')
}
```

```{r echo=FALSE, Multivariate_Plots}
plot.multi('alcohol','volatile.acidity')+
   ggtitle("volatile acidity vs alcohol, with quality as color")
```

```{r echo=FALSE}
plot.multi('alcohol','sulphates')+
  ggtitle("sulphates vs alcohol, with quality as color")
```
```{r echo=FALSE}
plot.multi('alcohol','citric.acid')+
  ggtitle("citric acid vs alcohol, with quality as color")
```

The 3 plot above demostrated that our findings in bivariable section:  
Alcohol, volatile acid has the medium correlation with quality, in their plot it is obvious that point cluster with range.  
Citric acid and sulphates has little correlation with quality, since in the point their point also cluster with range, but not obvious.
# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

The Multivariate Plots section didn't give us any new knowledge by visuliazing the data in a new way, it demostrated that our investigation on Bivariate section is sound.

By visuliazing the dataset in this way, we could see the positive correlation between quality and alcohol(quality 5 most falls in alcohol range 8-11, quality 6 most falls in range 10-12), negative correlation between quality and volatile acidity etc. 

In one word, visuliazing in this way strengthened our findings in bivariable section.


------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
ggplot(aes(x = quality), data = wine)  + 
geom_bar(aes(y = (..count..)/sum(..count..))) + 
ggtitle("Plot of quality distribution") + 
ylab('percentage distribution')+
  xlab("quality score")
```

### Description One

This plot shows the distribution of the overall quality in our dataset.

It tells us that the quality feature is heavily biased, with half of its range(3,4,8) take around only 5% sample and another  half (5,6,7) take around 95% sample of our dataset.

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(aes(x = factor(quality), y = alcohol), data = wine) +
  geom_boxplot()  +
  xlab("quality score") + ggtitle("Box plot of alcohol vs quality")+ 
  ylab("alcohol % by volume")
```

### Description Two

This plot shows the correlation between quality and alcohol.

Alcohol have the most correlation with quality in our dataset, and the plot shows the positive correlation between quality and alcohol clearly.

### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(aes(x = alcohol, y = volatile.acidity), data = wine) +
  geom_jitter(aes(color = factor(quality) )) + scale_color_brewer(type = 'seq')+
  ggtitle("volatile acidity vs alcohol, with quality as color")+
  xlab("alcohol % by volume")+
  ylab("volatile acidity in g/dm^3")
```

### Description Three

This plot shows distribution of quality with alcohol as x aixs and volatile acidity as y axis.

From this plot we could see a clearly positive correlation between quality and alcohol: most orange point(quality 5) on the left, yellow(quality 6) spread acrosss all the point, and dark blue point( quality 7) on the middle to right.

We could also see a nagetive correlation between volatile acidity and quality: Most orange point on the top,yellow on the middle and dark blue on the bottom.

------

# Reflection
The red wine dataset contains information on 1599 sample of wine, with 11 chemical feature describe its chemical characteristic and quality feature describe how good it is from peoples taste.

From the exploration we first plot every single feqature distribution. Then we have ggpair plot which plot every pair of feature in our dataset.  
From the ggpair plot we know some of the feature may have more chance to correlate with quality(alcohol: 0.476,volatile acidity:-0.391,sulphates:0.251 and citric acid:0.226 ).  
From bivariable plot we demostrated the correlation between quality and other feature, and multivariable plot strenthen our conclusion on the correation.

From the exploration and analysis we found that this dataset is highly biased:  
In quality feature, 95% of our sample within range (5,6,7), 82% of our sample are within range(5,6).  
Also, for the residual sugar feature, in the dataset variable description it says:"wines with greater than 45 grams/liter are considered sweet". But in the dataset, the highest value for this feature is around 16 g/liter, which means in our dataset no wine sample could be considered as sweet.  
Highly biased sampling make result of this dataset analysis not general. In another word, conclusion draw from this analysis could not apply to other red wine dataset.

Also since I'm not familiar with chemistry, I know really little about most of the feature in the dataset, thus may not able to dig deep into the dataset and find the story the data tell.

For the future work, I think we should have bigger and more balanced dataset(e.g. more sweet wine sample), thus we could make our analysis general. Also I think I could learn some field knowledge, to know more about how each feature could affect the wine quality.


